Factors affecting reliablity of state-space age-structured assessment models

### 2025-01-16 Project_0/manuscript/manuscript.Rmd

---
title: Factors affecting reliablity of state-space age-structured assessment models
author: 
  - Timothy J. Miller^1^
  - Andrew Applegate 
  - Greg Britten
  - Elizabeth N. Brooks
  - Gavin Fay
  - Alex Hansell
  - Christopher M. Legault
  - Brandon Muffley
  - Brian C. Stock
  - John Wiedenmann
date: ''
fontsize: 12pt
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    includes:
      in_header:
      - options.sty
      - journalshortcuts.tex
    keep_tex: yes
    number_sections: yes
  html_document:
    df_print: paged
  word_document: default
header-includes:
- \usepackage{xspace}
- \usepackage{bm}
- \usepackage{caption,graphics}
- \usepackage{graphicx}
- \usepackage{makecell}
- \renewcommand\figurename{Fig.}
- \captionsetup{labelsep=period, singlelinecheck=false}
- \newcommand{\changesize}[1]{\fontsize{#1pt}{#1pt}\selectfont}
- \renewcommand{\arraystretch}{1.5}
- \renewcommand\theadfont{}
csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
biblio-style: cjfas2.bst
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=FALSE, tidy.opts=list(width.cutoff=60), warning = FALSE, message = FALSE, fig.align = "center")
knitr::opts_knit$set(root.dir = '../' )
library(knitr)
library(tidyverse)
library(pander)
library(kableExtra)
library(png)
library(forcats)
```

$^1$corresponding author: timothy.j.miller@noaa.gov, Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\

@@

Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada, and the Northeast US \citep{nielsenberg14,cadigan16,pedersenberg17,stockmiller21}. State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models \citep{hoyleetal22, punt23}.

Much is known about the reliability of state-space models that are linear or Gaussian \citep{aeberhardetal18}, but applications in fisheries management are nonlinear and typically include multiple types of observations with varying distributional assumptions. Furthermore, there is a wide range of potential random effects structures and model parameters that can be treated as random effects in assessment models. We know relatively little about the factors affecting statistical reliability of such models or the ability of information criteria to distinguish among such alternative structures.

@@

We used the Woods Hole Assessment Model (WHAM) to configure operating and EMs in our simulation study \citep[]{millerstock20,stockmiller21}. WHAM is an R package freely available as a github repository. For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3). This packages has also been used to configure operating and EMs for closed loop simulations evaluating index-based assessment methods \citep{legaultetal23} and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, and black sea bass in the Northeast US.

We completed a simulation study with a number of OMs that can be categorized based on where process error random effects are assumed: abundance at age (R, R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q). For each OM assumptions about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from recent applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.

@@

We did not use the bias-correction feature for process errors or observations described by \citet{stockmiller21} for operating and EMs. Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster. All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code.

@@

The population model tracks 10 age classes: ages 1 to 10+ and we assume spawning occurs 1/4 of the way through the year. The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 and assumed known in all EMs (Figure \ref{om_maturity}).

Weight at age was generated with a LVB growth function
$$
L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
$$
where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a L-W relationship such that 
$$
W_a = \theta_1 L_a^{\theta_2}
$$
where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$ (Figure \ref{om_waa}).

We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters for all OMs. All post-recruit productivity components are constant in the NAA and survey catchability process error OMs. Therefore steepness and unfished recruitment are also constant over the time period for those OMs \citep{millerbrooks21}. We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = F_{40\%} = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
\[
N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
\]
Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without process errors on natural mortality we assumed the rate was assumed 0.2. For OMs with process errors on natural mortality the mean rate was 0.2.

Two alternative fishing histories were used for OMs. In the first scenario, the stock experiences overfishing for the first 20 years and fishing at \Fmsy for the last 20 years. In the second scenario, the stock is fished at \Fmsy for the entire time period. The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from \citet{wiedenmannetal19}. \Citet{legaultetal23} also used similar approaches to defining fishing mortality histories for OMs.

Initial population was configured at the equilibrium distribution fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$ for the two alternative fishing histories. That is for a deterministic model, the age composition would not change over time when the fishing mortality was constant at the respective level.

For OMs with time-varying random effects for M, steepness is not constant, but we used the same alpha and beta parameters as other OMs this equates to a steepness and R0 at the mean of the time series process for M. For OMs with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other OMs which corresponds to Fmsy at the mean selectivity parameters.

@@

Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 way through the year) and one in the fall (0.75 way through the year). Actually we have it currently configured that both occur 0.5 way through the year. Catchability of both surveys are assumed to be 0.1.  Like the fishing fleet, we assumed logistic selectivity for both indices with $a_{50} = 5$ and slope = 1 and a logistic-normal distribution for the age-composition observations.

@@

The factorial combinations of fishing histories, observation error assumptions, and marginal variance assumptions for recruitment and survival results in 24 different OMs with random effects on recruitment and survival (Table \ref{naa_om_table}).

@@

The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for natural mortality results in 16 different OMs with random effects on natural mortality (Table \ref{M_om_table}). Internally, WHAM treats natural mortality as a log-transformed parameter 
$$
\log M_y = \mu_M + \epsilon_{M,y}
$$
that is a linear combination of a mean $\mu_M$ and any annual random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$. The values for the marginal variance and correlation parameters assumed in the OMs are given in Table \ref{M_om_table}.

@@

The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for selectivity parameters results in 16 different OMs with random effects on selectivity (Table \ref{sel_om_table}). Internally, WHAM treats selectivity parameter $s$ as a logit-transformed parameter 
$$
\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
$$
that is a linear combination of a mean $\mu_s$ and any annual random effects marginally distributed as $\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)$ where the lower and upper bounds of the parameter ($l_s$ and $u_s$) can be specified by the user. All selectivity parameters are either $a_50$ or slope parameters and we assume bounds of 0 and 10 for all selectivity parameters for all operating and EMs. The values for the marginal variance and correlation parameters assumed in the OMs are given in Table \ref{sel_om_table}.

@@

The factorial combinations of fishing histories, observation error assumptions, and marginal variance and autocorrelation assumptions for catchability results in 16 different OMs with random effects on catchability (Table \ref{q_om_table}). Internally, WHAM treats catchability for index $i$ as a logit-transformed parameter 
$$
\log\left(\frac{q_{i,y}-l_{i}}{u_{i}-q_{i,y}}\right) = \mu_i + \epsilon_{i,y}
$$
that is a linear combination of a mean $\mu_i$ and any annual random effects marginally distributed as $\epsilon_{i,y} \sim \text{N}\left(0,\sigma_i^2\right)$ where the lower and upper bounds of the catchability ($l_i$ and $u_i$) can be specified by the user. Here we assume bounds of 0 and 1000 for all operating and EMs. For operating and EMs with process errors on catchability, the temporal variation is only assumed for the first index. The values for the marginal variance and correlation parameters assumed in the OMs are given in Table \ref{q_om_table}

@@

@@

1. Did the optimization routine (stats::nlminb) complete without error?
2. Did the stats::nlminb convergence flag = 0 indicate successful convergence?
3. Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
4. Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
5. Did TMB::sdreport provide all standard errors < 100?

The first convergence criterion assesses whether the model crashes. The third is a measure that assesses how flat the likelihood is at the optimized point of the likelihood surface. The fourth and fifth criteria are specific to the hessian-based standard error reporting provided by TMB. the TMB::sdreport function will sometimes return standard error estimates even when the calculated hessian is not invertible (fourth criterion). It will also provide large standard errors for some parameters that are not estimated well or are near bounds on the transformed scale that is of primary interest. For example, variance parameters for random effects can often be estimated near 0 when the fit suggests no variation in the random effects or, equivalently, that the model without random effects is better.

@@

We measured the frequency of correct model selection using marginal AIC. For a given OM we the set of models that were considered all made the same assumptions on whether or not to estimate M and whether a stock-recruit function is assumed.

We also measured the frequency of correctly selecting models with Beverton-Holt stock recruit function assumed over models without the stock-recruit function (null model). Because the proportions of simulations where the correct model was chosen for a given set of OM factors implied generally poor performance of AIC, we analyzed the effect of variation in SSB on the probability of lower AIC of the BH models. We fit binomial generalize linear models to the observed AIC choice for each data set as a function of log(SD(SSB)) and OM.

@@

We estimated bias as the median of the relative errors across all simulations for a given OM and EM combination.
$$
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
$$
of annual SSB and fully-selected fishing mortality for each EM and constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in \citet{millerhyun18} and \citet{stockmiller21}. We also estimated bias of stock-recruit parameters for EMs that assumed the Beverton-Holt stock recruit function and the mean natural mortality rate parameter when estimated.

@@

@@

AIC generally performs weakly in determining stock recruit relationships unless there is large contrast in SSB. Interestingly, \citet{milleretal16} found AIC to prefer a stock-recruit relationship for SNEMA yellowtail flounder where there was a large change in stock size estimated over time and flatfish are well observed by the bottom trawl indices relative to some other guilds such as semi-pelagic species. Although we did not compare models with alternative stock-recruit relationships, we do not expect AIC to perform any better distinguishing these relationships. Our results comport well with those of \citet{devalpinehastings02} where state-space stock-recruit models were fit just to SSB and recruitment "observations".

@@

\bibliography{paper}
<!-- \bibliography{project-0-paper}
\bibliographystyle{/home/tmiller2/work/bibliographies/cjfas2}  -->

@@

\begin{landscape}
\begin{table}
\caption{Distinguishing characteristics of the operating models with random effects on survival. Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices). Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant.}\label{naa_om_table}
{\footnotesize \input{naa_om_tab}}
\end{table}
\end{landscape}

\begin{landscape}
\begin{table}
\caption{Distinguishing characteristics of the operating models with random effects on natural mortality. Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices). Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant. For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{M_om_table}
{\input{M_om_tab}}
\end{table}
\end{landscape}

\begin{landscape}
\begin{table}
\caption{Distinguishing characteristics of the operating models with random effects on selectivity. Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices). Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant. For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{sel_om_table}
{\input{Sel_om_tab}}
\end{table}
\end{landscape}

\begin{landscape}
\begin{table}
\caption{Distinguishing characteristics of the operating models with random effects on catchability. Standard deviations (SD) are for log-normal distributed indices and logistic normal distributed age composition observations (fleet and indices). Fishing mortality changes after year 20 (of 40) for fishing histories where fishing mortality is not constant. For AR1 process errors, $\sigma$ is defined for the marginal distribution of the processes.}\label{q_om_table}
{\input{q_om_tab}}
\end{table}
\end{landscape}

\begin{table}
\caption{Distinguishing characteristics of the estimating models.}\label{em_table}
{\scriptsize \input{em_tab}}
\end{table}

<!-- AIC TABLES --->

\begin{table}
\caption{Operating models and estimation models all assume matching R or R+S process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{naa_om_em_R_BH_aic_table}
{\input{naa_om_em_R_BH_aic_table}}
\end{table}

\begin{table}
\caption{Operating models and estimation models all assume matching R+M process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{M_om_em_R_BH_aic_table}
{\input{M_om_em_R_BH_aic_table}}
\end{table}

\begin{table}
\caption{Operating models and estimation models all assume matching R+Sel process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{Sel_om_em_R_BH_aic_table}
{\input{Sel_om_em_R_BH_aic_table}}
\end{table}

\begin{table}
\caption{Operating models and estimation models all assume matching R+q process error structure, estimating models assume mean recruitment or a B-H stock recruit relationship and M is either fixed at the true value or estimated.}\label{q_om_em_R_BH_aic_table}
{\input{q_om_em_R_BH_aic_table}}
\end{table}

\clearpage
<!-- OM configuration plots --->

\begin{figure}
\caption{The proportion mature at age assumed for the population in all operating and estimating models.}\label{om_maturity}
\begin{center}
\includegraphics[width = \textwidth]{om_maturity.png}
\end{center}
\end{figure}

\begin{figure}
\caption{The weight at age assumed for the population in all operating and estimating models.}\label{om_waa}
\begin{center}
\includegraphics[width = \textwidth]{om_waa.png}
\end{center}
\end{figure}

\begin{figure}
\caption{The Beverton-Holt stock recruit relationship assume for all operating models.}\label{om_sr}
\begin{center}
\includegraphics[width = \textwidth]{om_sr.png}
\end{center}
\end{figure}

\begin{figure}
\caption{The selectivity at age assumed for the fishing fleet in operating models without selectivity random effects. The mean selectivity at age across time for the fishing fleet in operating models with selectivity random effects.}\label{om_mean_selectivity}
\begin{center}
\includegraphics[width = \textwidth]{om_mean_selectivity.png}
\end{center}
\end{figure}

<!-- Convergence --->

\begin{landscape}
\begin{figure}
\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S. vertical lines represent 95\% confidence intervals. All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{naa_om_em_R_MF_convergence}
\begin{center}
\includegraphics[width = \textwidth]{naa_om_p_convergence_meanR_M_fixed.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S. vertical lines represent 95\% confidence intervals. All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{naa_om_em_R_ME_convergence}
\begin{center}
\includegraphics[width = \textwidth]{naa_om_p_convergence_meanR_M_estimated.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S. vertical lines represent 95\% confidence intervals. All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{naa_om_em_BH_MF_convergence}
\begin{center}
\includegraphics[width = \textwidth]{naa_om_p_convergence_BH_M_fixed.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structures R and R+S. vertical lines represent 95\% confidence intervals. All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{naa_om_em_BH_ME_convergence}
\begin{center}
\includegraphics[width = \textwidth]{naa_om_p_convergence_BH_M_estimated.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M. vertical lines represent 95\% confidence intervals. All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{M_om_em_R_MF_convergence}
\begin{center}
\includegraphics[width = \textwidth]{M_om_p_convergence_meanR_M_fixed.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M. vertical lines represent 95\% confidence intervals. All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is estimated.}\label{M_om_em_R_ME_convergence}
\begin{center}
\includegraphics[width = \textwidth]{M_om_p_convergence_meanR_M_estimated.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M. vertical lines represent 95\% confidence intervals. All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is fixed at the true value.}\label{M_om_em_BH_MF_convergence}
\begin{center}
\includegraphics[width = \textwidth]{M_om_p_convergence_BH_M_fixed.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+M. vertical lines represent 95\% confidence intervals. All estimating models estimate a Beverton-Holt stock-recruit relationship and and M is estimated.}\label{M_om_em_BH_ME_convergence}
\begin{center}
\includegraphics[width = \textwidth]{M_om_p_convergence_BH_M_estimated.png}
\end{center}
\end{figure}
\end{landscape}

\begin{landscape}
\begin{figure}
\caption{Probability of each type of convergence of estimating models with alternative process error assumptions for operating models that have process error structure R+Sel. vertical lines represent 95\% confidence intervals. All estimating models estimate mean recruitment rather than a stock-recruit relationship and and M is fixed at the true value.}\label{Sel_om_em_R_MF_convergence}
\begin{center}
\includegraphics[width = \textwidth]{Sel_om_p_convergence_meanR_M_fixed.png}
\end{center}
\end{figure}
\end{landscape}

>>>>

---
title: Factors affecting reliablity of state-space age-structured assessment models
author: 
  - Timothy J. Miller^1,2^
  - Greg Britten^3^
  - Elizabeth N. Brooks^2^
  - Gavin Fay^4^
  - Alex Hansell^2^
  - Christopher M. Legault^2^
  - Brandon Muffley^5^
  - Brian C. Stock^6^
  - John Wiedenmann^7^
date: "`r format(Sys.time(), '%d %B, %Y')`"
fontsize: 12pt
output:
  pdf_document:
#    citation_package: natbib
    fig_caption: yes
    includes:
      in_header:
      - options.sty
      - journalshortcuts.tex
    keep_tex: yes
    number_sections: yes
  html_document:
    df_print: paged
#  word_document: default
header-includes:
#- \usepackage{makecell}
csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl
bibliography: manuscript.bib
#biblio-style: cjfas2.bst
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy=FALSE, tidy.opts=list(width.cutoff=60), warning = FALSE, message = FALSE, fig.align = "center")
#knitr::opts_knit$set(root.dir = '../' )
library(knitr)
library(tidyverse)
library(pander)
library(kableExtra)
library(png)
library(forcats)
type <- "latex"
if(knitr::is_latex_output()) type <- "latex"
if(knitr::is_html_output()) type <- "html"
```

$^1$corresponding author: timothy.j.miller@noaa.gov\
$^2$Northeast Fisheries Science Center, Woods Hole Laboratory, 166 Water Street, Woods Hole, MA 02543 USA\
$^3$Woods Hole Oceanographic Institution\
$^4$SMAST\
$^5$Institute of Marine Research\
$^6$Mid-Atlantic Fisheries Management Council\
$^7$Rutgers University\

@@

Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada, and the Northeast US [@nielsenberg14;@cadigan16;@pedersenberg17;@stockmiller21]. State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models [@hoyleetal22; @punt23].

Much is known about the reliability of state-space models that are linear or Gaussian [@aeberhardetal18], but applications in fisheries management are nonlinear and typically include multiple types of observations with varying distributional assumptions. Furthermore, there is a wide range of potential random effects structures and model parameters that can be treated as random effects in assessment models. We know relatively little about the factors affecting statistical reliability of such models or the ability of information criteria to distinguish among such alternative structures.

@lietal24 investigated some aspects of inferences for operating models with multiple sources of process error, but there are differences for this paper.

@@

We used the Woods Hole Assessment Model (WHAM) to configure operating and EMs in our simulation study [@millerstock20;@stockmiller21]. WHAM is an R package freely available as a github repository. For this study we used [version 1.0.6.9000, commit 77bbd94](https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3). This package has also been used to configure operating and EMs for closed loop simulations evaluating index-based assessment methods [@legaultetal23] and is used for management of haddock, butterfish, American plaice, bluefish, Atlantic cod, black sea bass, and yellowtail flounder in the Northeast US.

We completed a simulation study with a number of OMs that can be categorized based on where process error random effects are assumed: abundance at age (R, R+S), natural mortality (R+M), fleet selectivity (R+Sel), or index catchability (R+q). For each OM assumption about variance of process errors and observations are required and the values we used were based on a review of the range of estimates from applications of WHAM in management of stocks of haddock, butterfish, and American plaice in the NE US.

@@

We did not use the log-normal bias-correction feature for process errors or observations described by [@stockmiller21] for operating and EMs [@lietalinreview]. Simulations and model fitting were all carried out on the University of Massachusetts Green High-Performance Computing Cluster. All code we used to perform the simulation study and summarize results can be found at https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code.

@@

The population consists of 10 age classes: ages 1 to 10+ and we assume spawning occurs each year 1/4 of the way through the year. The maturity at age was a logistic curve with $a_{50}$ = 2.89 and slope = 0.88 (Figure \ref{om_maturity}).

Weight at age was generated with a von Bertalanffy growth function
$$
L_a = L_{\infty}\left(1 - e^{-k(a - t_0)}\right)
$$
where $t_0 = 0$, $L_\infty = 85$, and $k = 0.3$, and a L-W relationship such that 
$$
W_a = \theta_1 L_a^{\theta_2}
$$
where $\theta_1 = e^{-12.1}$ and $\theta_2 = 3.2$ (Figure \ref{om_waa}).

We assumed a Beverton-Holt stock recruit function with constant pre-recruit mortality parameters for all OMs. All post-recruit productivity components are constant in the NAA and survey catchability process error OMs. Therefore steepness and unfished recruitment are also constant over the time period for those OMs [@millerbrooks21]. We specified unfished recruitment = $R_0 = e^{10}$ and $\Fmsy = F_{40\%} = 0.348$ equated to a steepness of 0.69 and $\alpha=0.60$ and $\beta = 2.4 \times 10^{-5}$ for the
\[
N_{1,y} = \frac{\alpha \text{SSB}_{y-1}}{1 + \beta \text{SSB}_{y-1}} 
\]
Beverton-Holt parameterization (Figure \ref{om_sr}). For OMs without process errors on natural mortality we assumed the rate was assumed 0.2. For OMs with process errors on natural mortality the mean log natural mortality rate was $\log(0.2)$.

We used two fishing scenarios for OMs. In the first scenario, the stock experiences overfishing at 2.5\Fmsy for the first 20 years and fishing at \Fmsy for the last 20 years (denoted $2.5\Fmsy \rightarrow \Fmsy$). In the second scenario, the stock is fished at \Fmsy for the entire time period. The magnitude of the overfishing assumptions is based on average estimates of overfishing for NE groundfish stocks from [@wiedenmannetal19]. @legaultetal23 also used similar approaches to defining fishing mortality histories for OMs.

We specified initial population abundance at age at the equilibrium distribution fishing at either $F = 2.5\times \Fmsy$ or $F = \Fmsy$ for the two alternative fishing histories. This implies that, for a deterministic model, the abundance at age would not change from the first year to the next.

For OMs with time-varying random effects for M, steepness is not constant, but we used the same alpha and beta parameters as other OMs this equates to a steepness and R0 at the median of the time series process for M. For OMs with time-varying random effects for fishery selectivity, \Fmsy is also not constant however we use the same F history as other OMs which corresponds to Fmsy at the mean selectivity parameters.

@@

Two time series of surveys are assumed and observed in numbers rather than biomass for the entire 40 year period with one occurring in the spring (0.25 of each year) and one in the fall (0.75 of each year).  Catchability of both surveys are assumed to be 0.1.  Like the fishing fleet, we assumed logistic selectivity for both indices with $a_{50} = 5$ and slope = 1 and a logistic-normal distribution for the age-composition observations.

@@

For operating models with random effects on recruitment and(or) survival (R, R+S) we assumed marginal standard deviations for recruitment of $\sigma_R \in \{0.5,1.5\}$ and marginal standard deviations for older age classes of $\sigma_{2+} \in \{0,0.25, 0.5\}$. The full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 24 different R ($\sigma_{2+} = 0$) and R+S operating models (Table \ref{naa_om_table}).

@@

All R+M OMs treat natural mortality constant across age, but with annually varying random effects. WHAM treats natural mortality as a log-transformed parameter 
$$
\log M_{y,a} = \mu_{M} + \epsilon_{M,y}
$$
that is a linear combination of a mean that is constant across ages $\mu_{M} = \log(0.2)$ and any annual random effects marginally distributed as $\epsilon_{M,y} \sim \text{N}\left(0,\sigma_M^2\right)$. Uncorrelated random effects were also included on recruitment with $\sigma_R = 0.5$ (hence, R+M). The marginal standard deviations we assumed for log natural mortality random effects were $\sigma_M \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_M \in \{0,0.9\}$. The full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 16 different R+M OMs (Table \ref{M_om_table}).

@@

MORE SPECIFICS about correlation  of random effects? Both selectivity pars? just correlated by year?
WHAM treats selectivity parameter $s$ as a logit-transformed parameter 
$$
\log\left(\frac{p_{s,y}-l_{s}}{u_{s}-p_{s,y}}\right) = \mu_s + \epsilon_{s,y}
$$
that is a linear combination of a mean $\mu_s$ and any annual random effects marginally distributed as $\epsilon_{s,y} \sim \text{N}\left(0,\sigma_s^2\right)$ where the lower and upper bounds of the parameter ($l_s$ and $u_s$) can be specified by the user. All selectivity parameters are either $a_50$ or slope parameters and we assume bounds of 0 and 10 for all selectivity parameters for all operating and EMs. The marginal standard deviations we assumed for logit scale random effects were $\sigma_s \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_s \in \{0,0.9\}$.  Like R+M OMs,  the full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 16 different R+Sel OMs (Table \ref{sel_om_table}).

@@

Like selectivity parameters, WHAM treats catchability for an index $i$ as a logit-transformed parameter 
$$
\log\left(\frac{q_{i,y}-l_{i}}{u_{i}-q_{i,y}}\right) = \mu_i + \epsilon_{i,y}
$$
that is a linear combination of a mean $\mu_i$ and any annual random effects marginally distributed as $\epsilon_{i,y} \sim \text{N}\left(0,\sigma_i^2\right)$ where the lower and upper bounds of the catchability ($l_i$ and $u_i$) can be specified by the user. Here we assume bounds of 0 and 1000 for all operating and EMs. For operating and EMs with process errors on catchability, the temporal variation is only assumed for the first index. The marginal standard deviations we assumed for logit scale random effects were $\sigma_i \in \{0.1, 0.5\}$ and AR1 autocorrelation parameters of $\rho_i \in \{0,0.9\}$.  Like R+M and R+Sel OMs, the full factorial combination of these process error assumptions and fishing history (2 levels) and observation error (2 levels) scenarios described above results in 16 different R+q OMs (Table \ref{q_om_table}).

@@

maturity, weight at age, index and catch CVs, assumed known in all EMs

@@

1. Fit complete: Did the optimization routine (stats::nlminb) complete without error?
2. No flag: Did the stats::nlminb convergence flag = 0 indicate successful convergence?
3. Good gradient: Was the maximum absolute value of the gradient of the log-likelihood < $1\times10^{-6}$?
4. No SE NAs: Did TMB::sdreport provide non-NA values for all fixed effects standard errors?
5. No big SEs: Did TMB::sdreport provide all standard errors < 100?

The first convergence criterion assesses whether the model crashes. The third is a measure that assesses how flat the likelihood is at the optimized point of the likelihood surface. The fourth and fifth criteria are specific to the hessian-based standard error reporting provided by TMB. the TMB::sdreport function will sometimes return standard error estimates even when the calculated hessian is not invertible (fourth criterion). It will also provide large standard errors for some parameters that are not estimated well or are near bounds on the transformed scale that is of primary interest. For example, variance parameters for random effects can often be estimated near 0 when the fit suggests no variation in the random effects or, equivalently, that the model without random effects is better.  Note that Types 2 through 5 also require type 1, and type 5 requires type 4. We used the Clopper-Pearson exact method for constructing 95\% confidence intervals of the probabilities of convergence [@clopperpearson34;@thulin14].

@@

We estimated the probability of selection of each process error model structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given operating model, we compared AIC for EMs that all made the same assumptions about median/constant natural mortality (known or estimated) and stock recruitment model (Beverton-Holt or none).

We also estimated the probability of correctly selecting between models with Beverton-Holt stock recruit function assumed and models without the stock-recruit function (null model). We made these comparisons between models that otherwise assume the same process error structure as the operating model and both of the compared models either estimate median/constant natural mortality or assume it is known. Our preliminary inspections of the proportions of simulations where the correct recruitment model was chosen for a given set of OM factors indicated generally poor performance of AIC. Furthermore, it has been shown that estimation of stock-recruit parameters requires a wide range of SSB (References). Therefore, we fit logistic regression models to the indicator of Beverton-Holt models having lower AIC as a function of the log-standard deviation of the true log(SSB) (similar to the log of the coefficient of variation for SSB) for each simulation.

All results only condition on whether all of the compared estimating models completed the optimization process without failure. We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because optimization can correctly determine the the correct likelihood that would indicate poor convergence because variance parameters may be at the lower bound of zero correctly for models that assume the incorrect process error structure.

@@

We estimated bias as the median of the relative errors across all simulations for a given OM and EM combination.
$$
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
$$
of annual SSB and fully-selected fishing mortality for each EM and constructed 95\% confidence intervals for the median relative bias using the binomial distribution approach as in @millerhyun18 and @stockmiller21. We also estimated bias of stock-recruit parameters for EMs that assumed the Beverton-Holt stock recruit function and the mean natural mortality rate parameter when estimated.

Similar to the AIC results, bias results only condition on whether the estimating model completed the optimization process without failure. We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because the optimized model can provide reliable estimation of SSB, F, M, and stock recruit parameters whether or not the model was able to estimate non-zero random effects. In practice, the model would be reconfigured to remove unnecessary process errors and produce otherwise equivalent parameter estimates.

@@

Similar to the other results, retrospective results only condition on whether all of the peels of a given estimating model completed the optimization process without failure. We did not condition on convergence as defined by a gradient threshold or invertibity of the hessian because the optimized model can provide reliable estimation of SSB, F, M, and stock recruit parameters whether or not the model was able to estimate non-zero random effects.

@@

AIC generally performs weakly in determining stock recruit relationships unless there is large contrast in SSB. Interestingly, @milleretal16 found AIC to prefer a stock-recruit relationship for SNEMA yellowtail flounder where there was a large change in stock size estimated over time and flatfish are well observed by the bottom trawl indices relative to some other guilds such as semi-pelagic species. Although we did not compare models with alternative stock-recruit relationships, we do not expect AIC to perform any better distinguishing these relationships. Our results comport well with those of @devalpinehastings02 where state-space stock-recruit models were fit just to SSB and recruitment "observations".

@@

# References {-}

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width = \textwidth]{om_input_plots_figure}
\end{center}
\caption{The proportion mature at age, weight at age, fleet and index selectivity at age, and Beverton-Holt stock recruit relationship assumed for the population in all operating models. For operating models with random effects on fleet selectivity, this represents the selectivity at the mean of the random effects.}\label{om_inputs_fig}
\end{figure}
